Metric Spaces with Expensive Distances

### 2019-12-04 paper/draft_socg.tex
Add Lab Earth Mover data, move grey value to appendix

However, we argue that in some situations, distance computations
in $\metricspace$ can be costly and $\complexity$ might be incomparable
with $n$. Our motivation comes from topological summaries
such as persistence diagrams~\cite{elz-topological} or Reeb graphs~\cite{reeb-survey}, which are of interest
in the field of topological data analysis. A persistence diagram
is a point set in $\RR^2$, and the distance between two diagrams
is determined by a min-cost matching between the point sets.
If the diagrams have $N$ points, computing this matching requires
polynomial time in $N$, and $N$ might well be larger than $n$, the number
of diagrams considered \cite{cohen2007stability}. For the case of Reeb graphs, the situation is even
worse: while several metrics on Reeb graphs have been proposed (\cite{bauer2014measuring}, \cite{de2016categorified},
\cite{di2016edit}),
not even a constant-factor approximation algorithm is known that runs
in polynomial time in the size of the graphs.
Another instance is a collection of images
equipped with the Wasserstein (or the Earth Mover's) metric \cite{rubner2000earth}.

@@

Our experiments with the Earth Mover's distance and persistence
diagrams of 3-D shapes show that blind greedy spanner can speed up computations
by a factor of 2 to 5, saving hundreds of hours of machine time,
even though the inputs in our data sets are of moderate size.
%of persistence diagrams of moderate size, computing a blind greedy spanner is faster than (approximately)
%computing all $\binom{n}{2}$ distances.
Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
We also compare with a spanner construction based on WSPD. Our simple algorithm gives much smaller
spanners on the tested examples. 
Nevertheless, we leave the question open whether our construction
yields a spanner of asymptotically linear size.

@@

\ignore{
The last option that we consider is based on an additional assumption: while
computing $\dist(\cdot, \cdot)$ exactly is costly, we have access to a
cheap approximation algorithm
that gives a $2$-approximation of $\dist$ much faster 
(the output of this approximation algorithm is guaranteed to
lie in $[\dist(p,q), 2 \dist(p,q)]$). In order to simplify notation, we now assume
that all pairwise distances between our points belong to $[1, 2^t]$ for some $t$.
We introduce \textit{buckets} of the form $[2^k, 2^{k+2}]$ for $k = 0 \dots t$;
these intervals are not disjoint, but by calling the approximation algorithm we can assign
each pair $(p_i, p_j)$ to one of the buckets.
Indeed, let $v$ be the output of the approximation algorithm for $\dist(p,q)$,
then we know that the true distance is in $[v/2, v]$. We want to put the pair $(p,q)$ into the bucket
that fully contains this interval, thus we need to find $k$ such that
$2^k \leq v /2$ and $v \leq 2^{k+2}$. These inequalities give $ 2^{k+1} \leq v \leq 2^{k+2}$,
and we take $k = \lfloor \log v \rfloor - 1$.
Afer we found a bucket for each pair $(p_i, p_j)$, 
we have some sort of a weak ordering: we cannot say anything
about two edges that are assigned to the same bucket or to overlapping buckets, but if the buckets
are disjoint, we know which edge is longer. Now we can either imitate the greedy spanner (non-blind) algorithm,
traversing buckets in the order of their left endpoint, and adding edges from them; or we could alternate: in even iterations
we traverse the buckets from left to right, adding edges from each bucket one by one, and in odd iterations
we traverse the buckets from right to left. This alternative makes sense, because we need to have good lower bounds in
order to build a blind spanner, and good lower bounds can be obtained from the triangle inequality, if one of the edges 
in a path connecting $p_i$ and $p_j$ in $G$ is longer than the sum of the lengths of all other edges.
We refer to the former option as \bqsgrdy and to the latter one as \bqsshaker.
}

@@

WSPD spanners performed poorly in our experiments
on non-clustered data, while the plots in a previous experimental study \cite{farshi2009experimental}
show that WSPD spanners are very sparse, outperformed only by the greedy algorithm.
We implemented two versions of  WSPD: one for the Euclidean case,
using quadtrees and the algorithm from \cite{hp-book}, and WSPD for general metric
spaces with cover trees (using the base $\tau=1.3$). 
They both give similar results, and we can only conclude
that the advantage of WSPD shows up on larger point sets than the ones we deal with.
The paper \cite{farshi2009experimental} contains experiments for up to 30000 points,
and our blind algorithms, which have cubic complexity in the number of points,
are infeasible for such $n$.

@@

\begin{table}\centering
    \ra{1.2}
\begin{tabular}{@{}rrrcrrcrr@{}}\toprule
                          & \multicolumn{2}{c}{$q=1$} & \phantom{abc} & \multicolumn{2}{c}{$q=2$} & \phantom{abc} & \multicolumn{2}{c}{$q=3$}\\
                          \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
                          & Greedy & Blind greedy && Greedy & Blind greedy && Greedy & Blind greedy \\ \midrule
$\eps = 0.1$                & 0.14  & 0.28         && 0.32  & 0.72      &&   0.41  & 0.78 \\
$\eps = 0.2$                & 0.06 & 0.19          && 0.13 & 0.44       &&   0.20  & 0.52 \\
$\eps = 0.4$               & 0.03 &  0.13          && 0.05 & 0.26        &&   0.06  & 0.34 \\
$\eps = 0.5$               & 0.02 &  0.10          && 0.03 & 0.20        &&   0.05  & 0.27 \\
$\eps = 0.8$               & 0.01 & 0.08           && 0.02  & 0.13        &&   0.02  & 0.17 \\
$\eps = 1.0$               & 0.01 &  0.06  && 0.01 & 0.10        &&   0.02  & 0.14 \\
\bottomrule
\end{tabular}
    \caption{Sparseness of \bgrdy and \textsc{Greedy} spanners for 450 \dtype{original} persistence diagrams
 with Wasserstein distance $W_q$.}
\label{tbl:mcgill_original_blind_greedy_spanner_sparseness}
\end{table}

\begin{table}\centering
    \ra{1.2}
    \begin{tabular}{@{}rrrcrrcrr@{}}\toprule
                          & \multicolumn{2}{c}{$q=1$} & \phantom{c} & \multicolumn{2}{c}{$q=2$} & \phantom{c} & \multicolumn{2}{c}{$q=3$}\\
                          \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
                          & Brute-force & Blind greedy && Brute-force & Blind greedy && Brute-force & Blind greedy \\ \midrule
$\eps = 0.1$    &   $2605$  &  $1266$  &&  $3719$  &  $4085$  &&  $11$\,$781$  &  $13$\,$557$\\
$\eps = 0.2$    &   $2209$  &  $878$  &&  $3338$  &  $2574$  &&  $10$\,$541$  &  $9383$ \\
$\eps = 0.4$    &   $1988$  &  $586$  &&  $3088$  &  $1590$  &&  $-$    &  $6383$ \\
$\eps = 0.5$    &   $1744$  &  $483$  &&  $2835$  &  $1233$  &&  $9418$    &  $5146$ \\
$\eps = 0.8$    &   $1691$  &  $368$  &&  $2809$  &  $814$  &&  $-$    &  $3419$ \\
$\eps = 1.0$    &   $1645$  &  $324$  &&  $2780$  &  $672$  &&  $-$    &  $-$ \\
\bottomrule
\end{tabular}
    \caption{Time required to approximate all pairwise distances with \bgrdy algorithm.
    Data is for 450 \dtype{original} persistence diagrams with Wasserstein distance $W_q$, time in seconds.}
\label{tbl:mcgill_original_blind_greedy_spanner_timing}
\end{table}

\begin{table}\centering
    \ra{1.2}
    \begin{tabular}{@{}rrrcrrrc@{}}\toprule
        & \multicolumn{2}{c}{Sparseness} & \phantom{abc} & \multicolumn{3}{c}{Running time} \\
        \cmidrule{2-3} \cmidrule{5-7} 
                & Greedy  & Blind greedy && Brute-force & Blind greedy & Ratio \\ \midrule
$\eps = 0.2$    &   $0.34$  &  $0.81$    &&  $185\,863$   & $133\,194$    & $0.72$   \\
$\eps = 0.3$    &   $0.20$  &  $0.69$    &&  $185\,863$   & $105\,723$    & $0.56$   \\
$\eps = 0.4$    &   $0.13$  &  $0.58$    &&  $185\,863$   & $86\,283$    & $0.46$   \\
$\eps = 0.5$    &   $0.09$  &  $0.50$    &&  $185\,863$   & $72\,447$    & $0.38$   \\
$\eps = 0.8$    &   $0.05$  &  $0.35$    &&  $185\,863$   & $48\,870$    & $0.26$   \\
$\eps = 1.0$    &   $0.04$  &  $0.30$    &&  $185\,863$   & $40\,955$    & $0.22$   \\
\bottomrule
\end{tabular}
    \caption{Results of \bgrdy algorithm on the dataset of 200 images
    with Earth Mover's Distance on \textit{Lab} histograms, time in seconds.}
\label{tbl:emd_gtsrb}
\end{table}

The other data sets that we used
are the persistence diagrams
computed from the McGill shape benchmark \cite{zhang2005mcgill}.
More precisely, we generated two sets of diagrams,
\dtype{original} and \dtype{perturbed}. The \dtype{original}
diagrams were computed on the points of the 450 original
point clouds from the shape data set; we only used
persitence in dimension 0, since the diagrams in this dimension are larger in size.
For the \dtype{perturbed} data set,
we added random noise to points of the original shapes,
creating 10 modified versions of each of them;
then we computed persistence diagrams in dimension 0 of these modified
shapes.
Persistence diagrams were computed with the \libraryname{DIPHA}
library \cite{bauer2014distributed}. 
Wasserstein distance were computed approximately
with the software library \hera using a relative error of $0.01$.
We restricted to approximate computation because computing the
Wasserstein distance exactly (using the Hungarian algorithm)
has been reported to be very expensice~\cite{kerber2017geometry}.

Sparseness values of \bgrdy spanner on the whole \dtype{original} data set
for different values of $\eps$ and $q$ are presented in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness};
the \grdy algorithm is used as baseline.  We observe that for higher values of $q$
the advantage of our algorithm fades away. A plausible explanation
is that for higher values of $q$, the doubling dimension becomes larger;
this is confirmed by the increase in size of the greedy spanner.
For $\wsdist{1}$ we observe more encouraging figures, avoiding 70 \% of distance
computations for $\eps = 0.1$; for a $1.5$-approximation, we
reduce the number of distances computed by a factor of 10.

@@

\end{axis}
\end{tikzpicture}
\end{centering}
    \caption{Sparseness of \bgrdy spanner algorithm. The plot is for \dtype{perturbed} persistence diagrams,
    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
    \label{fig:spanner_sparseness_mcgill_perturbed}
\end{figure}

We also compare the time that is needed to compute the matrix of all pairwise distances
naively with the time that is needed to approximate it with \bgrdy spanner.
The timings are given in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
We compared our approach for a fixed
$\eps$ with the time to compute all pairwise-distance between the
persistence diagrams using \hera, using the same approximation parameter
$\eps$. Let us emphasize that the diagrams we have are rather small,
with average cardinality
350 points. Nevertheless, on this data set \bgrdy spanner
provides a substantial improvement in terms of real running time,
and we expect this advantage to grow for diagrams of larger cardinality.

In order to see the growth of the spanner size, we use the \dtype{perturbed}
data set, drawing from it random samples.
We plot the sparseness of the spanners obtained by \grdy
and \bgrdy algorithms in \cref{fig:spanner_sparseness_mcgill_perturbed}
for different values of $\eps$, always using the $\wsdist{1}$ distance.
As expected, the ratio of computed distances decreases as the number of diagrams
grows. In \cref{fig:spanner_ratio_mcgill_perturbed} we show
the growth of the spanner size relative to the number of diagrams.
For $\eps = 0.5$ the \bgrdy spanner starts to stabilize (i.e., demonstrates
linear growth) for larger values of $n$.

@@

\end{axis}
\end{tikzpicture}
\end{centering}
    \caption{Ratio \# edges /  \# diagrams for  \bgrdy spanner. The plot is for 0-dimensional diagrams of perturbed McGill data
    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
    \label{fig:spanner_ratio_mcgill_perturbed}
\end{figure}

For experiments with images, we used the German traffic sign
dataset GTSRB \cite{gtsrb} and the code for computing the Earth Mover's distance
by Doran \cite{doran_pyemd}.
We randomly selected 250 images from the benchmark.
The results of the experiment are summarized in \cref{tbl:emd_gtsrb}.
The first two columns compare the sparseness of \bgrdy and greedy spanners.
For smaller values of $\eps$,
the \bgrdy spanner contains approximately 3 times more
edges than the greedy spanner, so the ratio is roughly
the same as for persistence diagrams.
The last 3 columns show the advantage
of \bgrdy in terms of the running time.
If a $1.5$ approximation is acceptable,
then the \bgrdy algorithm speeds up
the computation by a factor of more than $5$,
and even for a $10 \%$ approximation, it reduces
the running time by more than a factor of $2$,
saving $98$ hours of computation time.

@@

Now, traversing the points $P$ in random order,
we call a point a \emph{minimum} if it is closer to $q$
than all previous points. Note that minima do not necessarily
correspond to the candidates in the algorithm because
we look for an approximate neareast neighbor only.
Standard backwards analysis~\cite{seidel-backwards}
shows that the expected number of minima encountered is $O(\log n)$.
The first two ideas together with a packing argument show
that between two consecutive minima, 
the algorithm queries only a constant number of distances,
and the bound follows.

@@

\section{Conclusion and future work}
\label{sec:conclusion}
%
We have introduced a new cost model for the analysis of algorithms
for metric spaces that fits the situation that computing an individual distance
is more costly than other types of primitive operations.
Our theoretical results assume
that the metric space has a low doubling dimension.
However, in our motivating example of collections of persistence diagrams
or Reeb graphs, this assumption does not hold. For instance,
the space of persistence diagrams has an infinite doubling dimension.
Nevertheless, realistic data sets are usually not just a random sample
in that infinite-dimensional space, but have structures
(e.g. clusters of close-by diagrams) which should be favourable for our approach.
Moreover, the size of the \bgrdy spanner can itself be considered as a piece
of information about the dimensionality of the data set under consideration.

@@

\section{Additional Experimental Results on Spanners}
\label{sec:spanners_euclidean_addendum}
In this section we include plots that give more information
about the behaviour of \bgrdy spanner in Euclidean space.

@@

@@

Let $I_1$, $I_2$ be two greyvalue images.
We divide each greyvalue of $I_{1,2}$
by the sum of all greyvalues in the corresponding image.
Thus we can interpret $I_i$
as a probability measure supported on a rectangle
that has one vertex at the origin and
fully lies in the positive quadrant $\{ (x, y) \| x \geq 0, y \geq 0 \}$.
The Earth Mover's distance between images $I_1$ and $I_2$
is, by definition, the Earth Mover's distance between
these probability measures.

>>>>

However, we argue that in some situations, distance computations
in $\metricspace$ can be costly and $\complexity$ might be incomparable
with $n$. Our motivation comes from topological summaries
such as persistence diagrams~\cite{elz-topological} or Reeb graphs~\cite{reeb-survey}, which are of interest
in the field of topological data analysis. A persistence diagram
is a point set in $\RR^2$, and the distance between two diagrams
is determined by a min-cost matching between the point sets.
If the diagrams have $N$ points, computing this matching requires
polynomial time in $N$, and $N$ might well be larger than $n$, the number
of diagrams considered \cite{cohen2007stability}. For the case of Reeb graphs, the situation is even
worse: while several metrics on Reeb graphs have been proposed (\cite{bauer2014measuring}, \cite{de2016categorified},
\cite{di2016edit}),
not even a constant-factor approximation algorithm is known that runs
in polynomial time in the size of the graphs.

Another example is the Earth Mover's distance (EMD, \cite{rubner2000earth}), which
is often used for images.
To compute it, one has to solve the optimal
transportation problem, which has cubical complexity. There exist many papers addressing this issue,
e.g., \cite{wemd}, \cite{grounded_emd}, \cite{entropy_emd}.
However, usually these solutions either provide no guarantees of the approximation
quality at all, as in entropy-regularization based approaches,
or give only a loose bound on the relative error (e.g., the wavelet EMD \cite{wemd}
theoretically can differ from the EMD by a factor of 7).

@@

Our experiments with the Earth Mover's distance and persistence
diagrams of 3-D shapes show that blind greedy spanner can speed up computations
by a factor of up to 5, saving hundreds of hours of machine time,
even though the inputs in our data sets are of moderate size.
%of persistence diagrams of moderate size, computing a blind greedy spanner is faster than (approximately)
%computing all $\binom{n}{2}$ distances.
Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
We also compare with a spanner construction based on WSPD. Our simple algorithm gives much smaller
spanners on the tested examples. 
Nevertheless, we leave the question open whether our construction
yields a spanner of asymptotically linear size.

@@

\ignore{
The last option that we consider is based on an additional assumption: while
computing $\dist(\cdot, \cdot)$ exactly is costly, we have access to a
cheap approximation algorithm
that gives a $2$-approximation of $\dist$ much faster 
(the output of this approximation algorithm is guaranteed to
lie in $[\dist(p,q), 2 \dist(p,q)]$). In order to simplify notation, we now assume
that all pairwise distances between our points belong to $[1, 2^t]$ for some $t$.
We introduce \textit{buckets} of the form $[2^k, 2^{k+2}]$ for $k = 0 \dots t$;
these intervals are not disjoint, but by calling the approximation algorithm we can assign
each pair $(p_i, p_j)$ to one of the buckets.
Indeed, let $v$ be the output of the approximation algorithm for $\dist(p,q)$,
then we know that the true distance is in $[v/2, v]$. We want to put the pair $(p,q)$ into the bucket
that fully contains this interval, thus we need to find $k$ such that
$2^k \leq v /2$ and $v \leq 2^{k+2}$. These inequalities give $ 2^{k+1} \leq v \leq 2^{k+2}$,
and we take $k = \lfloor \log v \rfloor - 1$.
After we found a bucket for each pair $(p_i, p_j)$, 
we have some sort of a weak ordering: we cannot say anything
about two edges that are assigned to the same bucket or to overlapping buckets, but if the buckets
are disjoint, we know which edge is longer. Now we can either imitate the greedy spanner (non-blind) algorithm,
traversing buckets in the order of their left endpoint, and adding edges from them; or we could alternate: in even iterations
we traverse the buckets from left to right, adding edges from each bucket one by one, and in odd iterations
we traverse the buckets from right to left. This alternative makes sense, because we need to have good lower bounds in
order to build a blind spanner, and good lower bounds can be obtained from the triangle inequality, if one of the edges 
in a path connecting $p_i$ and $p_j$ in $G$ is longer than the sum of the lengths of all other edges.
We refer to the former option as \bqsgrdy and to the latter one as \bqsshaker.
}

@@

WSPD spanners performed poorly in our experiments
on non-clustered data, while the plots in a previous experimental study \cite{farshi2009experimental}
show that WSPD spanners are very sparse, outperformed only by the greedy algorithm.
We implemented two versions of  WSPD: one for the Euclidean case,
using quad trees and the algorithm from \cite{hp-book}, and WSPD for general metric
spaces with cover trees (using the base $\tau=1.3$). 
They both give similar results, and we can only conclude
that the advantage of WSPD shows up on larger point sets than the ones we deal with.
The paper \cite{farshi2009experimental} contains experiments for up to 30000 points,
and our blind algorithms, which have cubic complexity in the number of points,
are infeasible for such $n$.

@@

\begin{table}\centering
    \ra{1.2}
\begin{tabular}{@{}rrrcrrcrr@{}}\toprule
                          & \multicolumn{2}{c}{$q=1$} & \phantom{a} & \multicolumn{2}{c}{$q=2$} & \phantom{a} & \multicolumn{2}{c}{$q=3$}\\
                          \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
                          & Greedy & Blind greedy && Greedy & Blind greedy && Greedy & Blind greedy \\ \midrule
$\eps = 0.1$                & 0.14  & 0.28         && 0.32  & 0.72      &&   0.41  & 0.78 \\
$\eps = 0.2$                & 0.06 & 0.19          && 0.13 & 0.44       &&   0.20  & 0.52 \\
$\eps = 0.4$               & 0.03 &  0.13          && 0.05 & 0.26        &&   0.06  & 0.34 \\
$\eps = 0.5$               & 0.02 &  0.10          && 0.03 & 0.20        &&   0.05  & 0.27 \\
$\eps = 0.8$               & 0.01 & 0.08           && 0.02  & 0.13        &&   0.02  & 0.17 \\
$\eps = 1.0$               & 0.01 &  0.06  && 0.01 & 0.10        &&   0.02  & 0.14 \\
\bottomrule
\end{tabular}
    \caption{Sparseness of the \bgrdy and the \textsc{Greedy} spanners for 450 \dtype{original} persistence diagrams
 with Wasserstein distance $W_q$.}
\label{tbl:mcgill_original_blind_greedy_spanner_sparseness}
\end{table}

\begin{table}\centering
    \ra{1.2}
    \begin{tabular}{@{}rrrcrrcrr@{}}\toprule
                          & \multicolumn{2}{c}{$q=1$} & \phantom{c} & \multicolumn{2}{c}{$q=2$} & \phantom{c} & \multicolumn{2}{c}{$q=3$}\\
                          \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
                          & BF & BG && BF & BG && BF & BG \\ \midrule
$\eps = 0.1$    &   $2605$  &  $1266$  &&  $3719$  &  $4085$  &&  $11\,781$  &  $13\,557$\\
$\eps = 0.2$    &   $2209$  &  $878$  &&  $3338$  &  $2574$  &&  $10\,541$  &  $9383$ \\
$\eps = 0.4$    &   $1988$  &  $586$  &&  $3088$  &  $1590$  &&  $10\,128$    &  $6383$ \\
$\eps = 0.5$    &   $1744$  &  $483$  &&  $2835$  &  $1233$  &&  $9418$    &  $5146$ \\
$\eps = 0.8$    &   $1691$  &  $368$  &&  $2809$  &  $814$  &&  $9578$    &  $3419$ \\
$\eps = 1.0$    &   $1645$  &  $324$  &&  $2780$  &  $672$  &&  $9539$    &  $2743$ \\
\bottomrule
\end{tabular}
    \caption{Time required to approximate all pairwise distances with the \bgrdy algorithm.
    Data is for 450 \dtype{original} persistence diagrams with Wasserstein distance $W_q$, time in seconds.
    BF stands for brute-force, BG for \bgrdy.}
\label{tbl:mcgill_original_blind_greedy_spanner_timing}
\end{table}

\myparagraph{Experiments with Persistence Diagrams}
The other data sets that we used
are the persistence diagrams
computed from the McGill shape benchmark \cite{zhang2005mcgill}.
More precisely, we generated two sets of diagrams,
\dtype{original} and \dtype{perturbed}. The \dtype{original}
diagrams were computed on the points of the 450 original
point clouds from the shape data set; we only used
persistence in dimension 0, since the diagrams in this dimension are larger in size.
For the \dtype{perturbed} data set,
we added random noise to points of the original shapes,
creating 10 modified versions of each of them;
then we computed persistence diagrams in dimension 0 of these modified
shapes.
Persistence diagrams were computed with the \libraryname{DIPHA}
library \cite{bauer2014distributed}. 
Wasserstein distance were computed approximately
with the software library \hera using a relative error of $0.01$.
We restricted to approximate computation, because computing the
Wasserstein distance exactly (using the Hungarian algorithm)
has been reported to be very expensive~\cite{kerber2017geometry}.

Sparseness values of the \bgrdy spanner on the whole \dtype{original} data set
for different values of $\eps$ and $q$ are presented in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness};
the \grdy algorithm is used as baseline.  We observe that for higher values of $q$
the advantage of our algorithm fades away. A plausible explanation
is that for higher values of $q$, the doubling dimension becomes larger;
this is confirmed by the increase in size of the greedy spanner.
For $\wsdist{1}$ we observe more encouraging figures, avoiding 70 \% of distance
computations for even $\eps = 0.1$; for a $1.5$-approximation, we
reduce the number of distances computed by a factor of 10.

@@

\end{axis}
\end{tikzpicture}
\end{centering}
    \caption{Sparseness of the \bgrdy spanner. The plot is for \dtype{perturbed} persistence diagrams,
    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
    \label{fig:spanner_sparseness_mcgill_perturbed}
\end{figure}

We also compare the time that is needed to compute the matrix of all pairwise distances
naively with the time that is needed to approximate it with the \bgrdy spanner.
The timings are given in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
We compare our approach with the time to compute all pairwise-distance between the
persistence diagrams with \hera, using the same approximation parameter
$\eps$ (that is the reason why the BF column contains different values)\footnote{Let us be completely clear. The time in the BF column is the total running time for computing 
$\wsdist{q}$ with the corresponding $\eps$, so \textit{all} distances in the BF case are approximate. The time in the BG column
is the total time of the \bgrdy algorithm, which internally computes $\wsdist{q}$ with $\eps = 0.01$. In particular,
the sparseness column in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness} shows the fraction of distances that are almost exact.}.
Let us emphasize that the diagrams we have are rather small,
with average cardinality of
350 points. Nevertheless, on this data set the \bgrdy spanner
provides a substantial improvement in terms of real running time,
and we expect this advantage to grow for diagrams of larger cardinality.

In order to see the growth of the spanner size, we use the \dtype{perturbed}
data set, drawing from it random samples.
We plot the sparseness of the spanners obtained by the \grdy
and the \bgrdy algorithms in \cref{fig:spanner_sparseness_mcgill_perturbed}
for different values of $\eps$, always using the $\wsdist{1}$ distance.
As expected, the ratio of computed distances decreases as the number of diagrams
grows. In \cref{fig:spanner_ratio_mcgill_perturbed} we show
the growth of the spanner size relative to the number of diagrams.
For $\eps = 0.5$ the \bgrdy spanner starts to stabilize (i.e., demonstrates
linear growth) for larger values of $n$.

@@

\end{axis}
\end{tikzpicture}
\end{centering}
    \caption{Ratio \# edges /  \# diagrams for the \bgrdy spanner. The plot is for 0-dimensional diagrams of perturbed McGill data
    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
    \label{fig:spanner_ratio_mcgill_perturbed}
\end{figure}

\begin{table}\centering
    \ra{1.2}
    \begin{tabular}{@{}rrrcrrrc@{}}\toprule
        & \multicolumn{2}{c}{Sparseness} & \phantom{abc} & \multicolumn{3}{c}{Running time} \\
        \cmidrule{2-3} \cmidrule{5-7} 
                & Greedy  & Blind greedy && Brute-force & Blind greedy & Ratio \\ \midrule
$\eps = 0.2$    &   $0.34$  &  $0.81$    &&  $185\,863$   & $133\,194$    & $0.72$   \\
$\eps = 0.3$    &   $0.20$  &  $0.69$    &&  $185\,863$   & $105\,723$    & $0.56$   \\
$\eps = 0.4$    &   $0.13$  &  $0.58$    &&  $185\,863$   & $86\,283$    & $0.46$   \\
$\eps = 0.5$    &   $0.09$  &  $0.50$    &&  $185\,863$   & $72\,447$    & $0.38$   \\
$\eps = 0.8$    &   $0.05$  &  $0.35$    &&  $185\,863$   & $48\,870$    & $0.26$   \\
$\eps = 1.0$    &   $0.04$  &  $0.30$    &&  $185\,863$   & $40\,955$    & $0.22$   \\
\bottomrule
\end{tabular}
    \caption{Results of the \bgrdy algorithm on the dataset of 200 images
    with Earth Mover's Distance on \textit{Lab} histograms, time in seconds.}
\label{tbl:emd_lab_caltech}
\end{table}

\myparagraph{Earth Mover's Distance} 
We took the standard approach \cite{rubner2000earth} of transforming a colored image
to the \textit{Lab} color space and creating a $16\times 16\times 16$
distribution histogram. After that we compute the distance between normalized
distributions with \cite{doran_pyemd}.
We used the Caltech 101 dataset \cite{caltech101},
randomly selecting 200 images from different (randomly chosen)
categories.  We measured only the time needed to compute
the Earth Mover's distance, without preprocessing (color space transformation
and binning). The results are presented in \cref{tbl:emd_lab_caltech}.
The first two columns of the table compare the sparseness of the \bgrdy and the greedy spanners.
The last 3 columns show the advantage
of \bgrdy in terms of the running time.
Even for the small relative error $\eps = 0.2$,
the \bgrdy spanner saves about $28\%$ of time,
and, if {$2$-approximation} is acceptable,
then the \bgrdy algorithm speeds up
the computation by a factor of almost $5$.

Alternatively, one could also compute the Earth Mover's distance
on gray value images by considering each image as a distribution on pixels.
The results of this experiment are in \cref{sec:spanners_euclidean_addendum}, since this variant
does not seem to be used in applications.

@@

Now, traversing the points $P$ in random order,
we call a point a \emph{minimum} if it is closer to $q$
than all previous points. Note that minima do not necessarily
correspond to the candidates in the algorithm because
we look for an approximate nearest neighbor only.
Standard backwards analysis~\cite{seidel-backwards}
shows that the expected number of minima encountered is $O(\log n)$.
The first two ideas together with a packing argument show
that between two consecutive minima, 
the algorithm queries only a constant number of distances,
and the bound follows.

@@

\section{Conclusion and future work}
\label{sec:conclusion}
%
We have introduced a new cost model for the analysis of algorithms
for metric spaces that fits the situation that computing an individual distance
is more costly than other types of primitive operations.
Our theoretical results assume
that the metric space has a low doubling dimension.
However, in our motivating example of collections of persistence diagrams
or Reeb graphs, this assumption does not hold. For instance,
the space of persistence diagrams has an infinite doubling dimension.
Nevertheless, realistic data sets are usually not just a random sample
in that infinite-dimensional space, but have structures
(e.g. clusters of close-by diagrams) which should be favorable for our approach.
Moreover, the size of the \bgrdy spanner can itself be considered as a piece
of information about the dimensionality of the data set under consideration.

@@

\section{Additional Experimental Results on Spanners}
\label{sec:spanners_euclidean_addendum}
In this section we include plots that give more information
about the behavior of \bgrdy spanner in Euclidean space.

@@

\myparagraph{Earth Mover's Distance on Images}
Let $I_1$, $I_2$ be two gray value images.
If we divide each gray value of $I_{i}$
by the total sum, we can interpret $I_i$
as a probability measure supported on a rectangle
that has one vertex at the origin and
fully lies in the positive quadrant $\{ (x, y) \| x \geq 0, y \geq 0 \}$.
The Earth Mover's distance between images $I_1$ and $I_2$
is, by definition, the Earth Mover's distance between
these probability measures.
Note that this defines a true metric on the space of images,
while the standard approach with color histograms only defines
a pseudometric, since any permutaion of pixels does not
change the distribution in the color space.
However, this metric is very hard to compute:
for images with resolution $40\times 40$,
it takes about 20 minutes. For this reason,
we cannot test it on the Caltech 101 dataset.
Instead, we used the German traffic sign
dataset GTSRB \cite{gtsrb}. This
dataset contains small images, and usually
one traffic sign occupies most of the image area.
It is interesting to see how
the \bgrdy spanner would behave in this situation.

We randomly selected 250 images from the dataset.
The results of the experiment are summarized in \cref{tbl:emd_gtsrb}.
As in \cref{sec:experiment_spanners}, 
the first two columns give the sparseness.
For smaller values of $\eps$,
the \bgrdy spanner contains approximately 3 times more
edges than the greedy spanner, so the ratio is roughly
the same as for persistence diagrams.
The last 3 columns compare the running time.
Note that even for $\eps=0.1$,
the \bgrdy spanner saves $70\%$ of the computation time.
If we allow large relative error,
we see the speed by almost a factor of $10$.

@@
